
```{r message = FALSE}
# Set the CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# load the required packages and install them if they are not.
source(here::here("code","setup.R"))

# getiing the working directory
wd <- here::here()

# Loading the data
crypto_data <- read.csv(here("data", "crypto_data.csv"))

# scaling factor
scaling_factor <- 100000
```

# Part 1: Financial returns and normality

The working directory is set to: `r wd`

## a) Load Bitcoin data and assess price stationarity

::: {.callout-note icon=false}

## Question

Read in the Bitcoin data from file Crypto data.csv. Then, assess the stationarity of the (raw)
Bitcoin prices.

:::

First, let's take a look at the Bitcoin Prices on a plot.


```{r warning = FALSE}
# Extract the Bitcoin prices
bitcoin_prices <- crypto_data$Bitcoin

# plot the Bitcoin prices
plot(bitcoin_prices,
     type="l",
     col="red", 
     main="Bitcoin Prices",
     xlab="Time",
     ylab="Price")
```

The graph of the raw Bitcoin prices suggest that the series might not be stationary.

Let's perform the Augmented Dickey-Fuller test to check if the raw Bitcoin prices are stationary.

```{r}
# test for stationarity
adf.test(crypto_data$Bitcoin)
```

Since the p-value is significantly bigger than 0.05, we can not reject the null hypothesis and therefore, we can conclude that the raw Bitcoin prices are non-stationary.

## b) Create and plot Bitcoin negative log returns, assess stationarity

::: {.callout-note icon=false}

## Question

Create a function to transform the Bitcoin prices into their negative log returns counterparts. Plot
the latter series and assess their stationarity. To compare the series, also plot the negative log
returns on a common scale.

:::

Let's create a function to compute the negative log returns of a given price series. We will then apply this function to the Bitcoin prices to compute the negative log returns.

```{r warning = FALSE}
# Function to compute negative log returns
negative_log_returns <- function(prices) {
  return(-diff(log(prices)))
}

# Compute negative log returns for Bitcoin prices
neg_log_returns_bitcoin <- negative_log_returns(bitcoin_prices)
```

We can now plot the negative log returns series and the raw Bitcoin prices to compare.

```{r warning = FALSE}
# Plot the negative log returns series
plot(neg_log_returns_bitcoin,
     type="l",
     col="blue", 
     main="Negative Log Returns of Bitcoin Prices", 
     xlab="Time",
     ylab="Negative Log Returns")
```

If we scale our data, we have the following plot:

```{r warning = FALSE}
# Remove the first price to match the length of neg_log_returns_bitcoin
trimmed_bitcoin_prices <- bitcoin_prices[-1]

# scale the negative log returns and the bitcoin prices
scaled_b <- data.frame(neg_log_returns_bitcoin = neg_log_returns_bitcoin,
                       trimmed_bitcoin_prices = trimmed_bitcoin_prices)
scaled_b$scaled_neg_log_returns_bitcoin <- scale(scaled_b$neg_log_returns_bitcoin)
scaled_b$scaled_trimmed_bitcoin_prices <- scale(scaled_b$trimmed_bitcoin_prices)

# Plot the scaled negative log returns
plot(scaled_b$scaled_neg_log_returns_bitcoin,
     type = "l",
     col = "blue",
     ylab = "Scaled Values",
     xlab = "Index", 
     main = "Scaled Negative Log Returns and Bitcoin Prices",
     ylim = range(c(scaled_b$scaled_neg_log_returns_bitcoin,
                    scaled_b$scaled_trimmed_bitcoin_prices)))

# Add the scaled bitcoin prices to the same plot
lines(scaled_b$scaled_trimmed_bitcoin_prices,
      col = "red")

# Add a legend to differentiate the two lines
legend("topright",
       legend = c("Scaled Neg Log Returns",
                  "Scaled Bitcoin Prices"),
       col = c("blue",
               "red"),
       lty = 1)
```

Visually, the negative log returns series does not appear to indicate a clear trend or seasonality. The variance, although it fluctuates in the middle, seems relatively constant. This observation suggests that the series may be stationary. To confirm this, we will perform the Augmented Dickey-Fuller test to assess the stationarity of the negative log returns.

```{r warning = FALSE}
# Perform the Augmented Dickey-Fuller test on negative log returns
adf_test_neg_log <- adf.test(neg_log_returns_bitcoin)

# Display the test results
print(adf_test_neg_log)
```

Since the p-value is significantly smaller than 0.05, we can reject the null hypothesis and conclude that the negative log returns series is stationary.

## c) Check negative log returns normality with histograms, QQ-plots, Anderson-Darling

::: {.callout-note icon=false}

## Question

Are the negative log returns normally distributed? Draw histograms, check QQ-plots and use an
Anderson-Darling testing procedure to answer this question.

:::

Let's first plot the histogram and QQ-plot of the negative log returns to visually assess the normality.

```{r}
# Plot histogram of the negative log returns
hist(neg_log_returns_bitcoin,
     breaks=50,
     col="lightblue", 
     main="Histogram of Negative Log Returns",
     xlab="Negative Log Returns")

# Plot QQ-plot of the negative log returns
qqnorm(neg_log_returns_bitcoin)
qqline(neg_log_returns_bitcoin,
       col="red")

# Perform Anderson-Darling test for normality
ad_test <- ad.test(neg_log_returns_bitcoin)

# Display the Anderson-Darling test result
print(ad_test)
```

Even though the Histogram suggest that the negative log returns follows a normal distribution, the p-value when performing the Andersen-Darling test is smaller than 5%. It indicates that the data does not follow a normal distribution. The Normal Q-Q plot suggest also that the data does not follow a normal distribution.

## d) Fit t-distribution, compare with Normal via QQ-plot analysis

::: {.callout-note icon=false}

## Question

Fit a t-distribution to the negative log returns using `fitdistr()`. Using a QQ-plot, decide whether the fit is better than with a Normal distribution, based on your answer in [(c)](#c-check-negative-log-returns-normality-with-histograms-qq-plots-anderson-darling).

:::

```{r warning = FALSE}
# Fit a t-distribution to the negative log returns
fit_t <- fitdistr(scaling_factor * neg_log_returns_bitcoin, "t") # Multiply by 100000 to avoid numerical issues

# Display the t-distribution fit parameters
print(fit_t)
```

The t-distribution fit parameters are: `r round(fit_t$estimate[1], 2)` for the mean, `r round(fit_t$estimate[2], 2)` for the standard deviation, and `r round(fit_t$estimate[3], 2)` for the degrees of freedom. We can now compare the QQ-plot of the t-distribution with the QQ-plot of the normal distribution of [question c)](#c-check-negative-log-returns-normality-with-histograms-qq-plots-anderson-darling).

```{r warning = FALSE}
par(mfrow = c(1, 2))

# Generate QQ-plot for t-distribution
df_t <- fit_t$estimate[3]  # Degrees of freedom from the fit
qqplot(rt(length(neg_log_returns_bitcoin),
          df=df_t),
       neg_log_returns_bitcoin,
       main="QQ-plot for t-distribution",
       xlab="Theoretical Quantiles (t-distribution)",
       ylab="Sample Quantiles")
qqline(neg_log_returns_bitcoin,
       col="red")

# Generate QQ-plot for normal distribution
qqnorm(neg_log_returns_bitcoin,
       main="QQ-plot for Normal distribution")
qqline(neg_log_returns_bitcoin,
       col="blue")

# Reset the graphic layout back to default
par(mfrow = c(1, 1))
```

As we can see, the QQ-plot for the t-distribution is closer to the 45-degree line than the QQ-plot for the normal distribution. This suggests that the t-distribution is a better fit for the negative log returns than the normal distribution.

## e) Compare t-distribution and normal tails

::: {.callout-note icon=false}

## Question

Compare the tails of the density of the t-distribution and the normal distribution. Can we expect more extreme, unexpected events in t-distribution or in normal distribution? What can you conclude about the extreme events of our bitcoin data?

:::


```{r}
# Fit a normal distribution to the unscaled negative log returns
fit_norm <- fitdistr(scaling_factor * neg_log_returns_bitcoin, "normal")

# Generate a sequence of values for the x-axis (log returns)
x <- seq(min(neg_log_returns_bitcoin), max(neg_log_returns_bitcoin), length = 100)

# Scale back the mean for plotting (for both normal and t-distributions)
# For normal distribution
scaled_mean_norm <- fit_norm$estimate[1] / scaling_factor
scaled_sd_norm <- fit_norm$estimate[2] / scaling_factor

# For t-distribution (mean and sd, degrees of freedom remains the same)
scaled_mean_t <- fit_t$estimate[2] / scaling_factor
scaled_sd_t <- fit_t$estimate[3] / scaling_factor

# Density for the normal distribution using the scaled mean and sd
dens_norm <- dnorm(x, mean = scaled_mean_norm, sd = scaled_sd_norm)

# Density for the t-distribution using the scaled parameters
dens_t <- dt((x - scaled_mean_t) / scaled_sd_t, df = fit_t$estimate[1]) / scaled_sd_t

# Plot the histogram of negative log returns
hist(neg_log_returns_bitcoin,
     breaks = 50,
     col = "lightblue",
     freq = FALSE,  # For density plot
     main = "Negative Log Returns with Fitted Distributions",
     xlab = "Negative Log Returns")

# Add the normal distribution curve
lines(x,
      dens_norm,
      col = "black",
      lwd = 2,
      lty = 1)

# Add the t-distribution curve
lines(x,
      dens_t,
      col = "red",
      lwd = 2,
      lty = 2)

# Add a legend
legend("topright",
       legend = c("Normal Distribution", "t-Distribution"),
       col = c("black", "red"),
       lty = c(1, 2),
       lwd = 2)
```

Visually, the tails of the t-distribution are heavier than those of the normal distribution. This means that the t-distribution assigns more probability to extreme events than the normal distribution. Therefore, we can expect more extreme, unexpected events in the t-distribution than in the normal distribution. This observation is consistent with the QQ-plot analysis in [question d)](#d-fit-t-distribution-compare-with-normal-via-qq-plot-analysis), where the t-distribution was a better fit for the negative log returns than the normal distribution.
