```{r  echo = FALSE, message = FALSE}
# load the required packages and install them if they are not.
source(here::here("code","setup.R"))

# Loading the data
crypto_data <- read.csv(here("data", "crypto_data.csv"))

# Parameters
rounding_digits <- 4
```

# Part 3: Financial returns and normality

## a) Check Bitcoin-ETH dependence using correlation

::: {.callout-note icon="false"}
## Question

Are the negative log returns of Bitcoin and ETH dependent? Compute the correlation using `cor.test()` function. Can we conclude that these series are independent?
:::

To answer this question, we will first plot both negative log returns and then perform a correlation test.

```{r echo = FALSE}
# Negative log returns function
negative_log_returns <- function(prices) {
  -diff(log(prices))
}

# Compute negative log returns for Bitcoin and Ethereum
neglogret_eth <- negative_log_returns(crypto_data$Ethereum)
neglogret_btc <- negative_log_returns(crypto_data$Bitcoin)

# Create a sequence for time indices matching the length of negative log returns
time_indices <- seq_along(neglogret_eth)

# Create a data frame with time indices and negative log returns
nlr_df <- data.frame(
  Time = time_indices,
  Ethereum = neglogret_eth,
  Bitcoin = neglogret_btc
)

# Reshape data to long format for plotting
nlr_long_df <- nlr_df %>%
  pivot_longer(
    cols = c(Ethereum, Bitcoin),
    names_to = "Crypto",
    values_to = "NegativeLogReturn"
  )

# Plot the negative log returns of Bitcoin and Ethereum
ggplot(nlr_long_df, aes(x = Time, y = NegativeLogReturn, color = Crypto)) +
  geom_line() +
  labs(
    title = "Negative Log Returns of Bitcoin and Ethereum",
    x = "Time (Minutes)",
    y = "Negative Log Return"
  ) +
  theme_minimal()
```

The graph could suggest that the negative log returns of Bitcoin and Ethereum are dependent. However, we will perform a correlation test to confirm this.

```{r echo = FALSE}
# Perform correlation test between Ethereum and Bitcoin negative log returns
cor_test_result <- cor.test(nlr_df$Ethereum, nlr_df$Bitcoin)
```

The computed correlation between the negative log returns of Bitcoin and Ethereum using the `cor.test()` function is `r round(cor_test_result$estimate, rounding_digits)` and is not statistically significant (p-value = `r round(cor_test_result$p.value, rounding_digits)`). However, the lack of significance in the correlation test does not necessarily imply independence, as there may be a non-linear relationship or other factors influencing the association between the two variables.

## b) Analyze Bitcoin-ETH cross-correlation function

::: {.callout-note icon="false"}
## Question

Calculate the cross-correlation function (CCF) between the negative log returns of Bitcoin and
ETH. What do you observe?
:::

```{r echo = FALSE}
# Calculate and plot the cross-correlation function (CCF)
ccf(nlr_df$Ethereum, nlr_df$Bitcoin, main = "Cross-Correlation Function: Ethereum vs. Bitcoin Negative Log Returns")
```

We can clearly observe a significant spike at lag 5 indicates that there is a notable correlation between Bitcoin’s returns and Ethereum’s returns when Ethereum’s returns are shifted by 5 minutes. Ethereum’s returns are correlated with past values of Bitcoin’s returns and thus may predict Etherum's returns with a 5 minutes delay. The spike indicates the strength of the correlation. In our case, the spike at lag 5 crosses the significance threshold (horizontal dashed lines). We can say with confidence that it is statistically significant and not likely due to random noise.
However, it is important to keep in mind that cross-correlation does not imply causation. Other factors or indirect relationships might also explain this pattern.

## c) Assess predictive power with Granger test

::: {.callout-note icon="false"}
## Question

Is one of the time series good predictor of the second? Assess whether there is any predictive power
between the negative log returns of Bitcoin and ETH. You can use `grangertest()` in the `lmtest`
package with carefully chosen hyperparameter `order`. What is your conclusion?
:::

```{r echo = FALSE}
# Testing if Ethereum Granger-causes Bitcoin
granger_test_eth_to_btc <- grangertest(Bitcoin ~ Ethereum, order = 5, data = nlr_df)

# Testing if Bitcoin Granger-causes Ethereum
granger_test_btc_to_eth <- grangertest(Ethereum ~ Bitcoin, order = 5, data = nlr_df)
```

We performed here two grander test using the `grandertest()`function. The first tests if Ethereum’s past values do not Granger-cause Bitcoin’s returns. With a p-value of `r round(granger_test_eth_to_btc$"Pr(>F)"[2],rounding_digits)`, we cannot reject the null hypothesis. Therefore, there is no evidence to suggest that Ethereum’s past values have predictive power for Bitcoin’s returns.

The second test examines if Bitcoin’s past values Granger-cause Ethereum’s returns. With a p-value of `r round(granger_test_btc_to_eth$"Pr(>F)"[2], rounding_digits)`, Bitcoin’s past values significantly Granger-cause Ethereum’s returns, meaning there is strong evidence of predictive power.

## d) Predict mutual impacts of drops

::: {.callout-note icon="false"}
## Question

Based on your answer in [(c)](#c-assess-predictive-power-with-granger-test), answer the following questions:
:::

### d.1) Predict ETH reaction to Bitcoin drop

::: {.callout-note icon="false"}
## Question

We observe an extreme sudden drop in Bitcoin stocks. What should we expect that will happen
with ETH stocks?
:::

Given our results in [question c)](#c-check-negative-log-returns-normality-with-histograms-qq-plots-anderson-darling), the sudden drop in Bitcoin stocks should have a significant impact on the Etherum stocks price. Given our analysis, the Ethereum stocks should drop 5 minutes (lag of 5 / [question b)](#b-analyze-bitcoin-eth-cross-correlation-function)) after the Bitcoin stocks drop with a value close to 80% of the Bitcoin drop.

### d.2) Predict Bitcoin reaction to ETH drop

::: {.callout-note icon="false"}
## Question

We observe an extreme sudden drop in ETH stocks. What should we expect that will happen
with Bitcoin stocks?
:::

Based on the Granger causality test results from [question c)](#c-check-negative-log-returns-normality-with-histograms-qq-plots-anderson-darling), Ethereum’s past values do not Granger-cause Bitcoin’s returns, suggesting no significant predictive relationship. If Ethereum experiences an extreme sudden drop, it is less likely that Bitcoin will exhibit a direct or immediate reaction based on historical relationships. However, Bitcoin and Ethereum share common market factors and investor sentiment, so an indirect reaction (e.g., due to market-wide panic) is still possible, though it is not strongly supported by the causality analysis.


